### 2016-02-15 paper/1-introduction.tex
Made some small changes to fit everything nicely on 8 pages

But with the ever increasing complexity of simulation models, parallel simulation becomes necessary to perform the simulation within reasonable time bounds.
And while \textsf{Parallel DEVS}~\cite{ParallelDEVS} was introduced to increase parallelism, this is often insufficient.
Several synchronization protocols from the discrete event simulation community~\cite{FujimotoBook} have been applied to \textsf{DEVS} simulation.
While several parallel \textsf{DEVS} simulation kernels exist, they are often limited to a single synchronization protocol.
The reason for different synchronization protocols, however, is that their distinct nature makes them applicable in different situations, each outperforming the other in specific models.
The applicability of parallel simulation capabilities of current tools is therefore limited.
% Paragraph moved to beneath intro dxex, else we can't refer to our tool
This paper introduces DEVS-Ex-Machina (``dxex''), our simulation tool which offers multiple synchronization protocols: no synchronization (sequential execution), conservative synchronization, or optimistic synchronization.
The selected synchronization protocol is transparent to the simulated model.
Users should merely determine, at the start of simulation, which protocol they wish to use.
Our tool is based on PythonPDEVS, but implemented in C++11, to increase both performance and portability across different platforms. Unlike PythonPDEVS dxex only supports multicore parallelism.

Users who simulate a wide variety of models, with different ideal synchronization protocols, are therefore in luck:
with our tool the model implementation can be reused with a different protocol to test which offers better performance.

@@

Dxex can be found online at \url{https://bitbucket.org/bcardoen/devs-ex-machina}.

The remainder of this paper is organized as follows:
Section~\ref{sec:2-background} introduces the necessary background on synchronization protocols.
Section~\ref{sec:3-features} elaborates on some of our features, and our design that enables our flexibility.
In Section~\ref{sec:4-performance}, we evaluate performance of our tool by comparing its different synchronization protocols, as well as a comparison to adevs.
Related work is discussed in Section~\ref{sec:5-related-work}.
Section~\ref{sec:6-conclusion} concludes the paper and gives future work.

>>>>

But with the ever increasing complexity of simulation models, parallel simulation becomes necessary to perform the simulation within reasonable time bounds.
And while \textsf{Parallel DEVS}~\cite{ParallelDEVS} was introduced to increase parallelism, this is often insufficient.
Several synchronization protocols from the discrete event simulation community~\cite{FujimotoBook} have been applied to \textsf{DEVS} simulation.
While several parallel \textsf{DEVS} simulation kernels exist, they are often limited to a single synchronization protocol.
The reason for different synchronization protocols, however, is that their distinct nature makes them applicable in different situations, each outperforming the other in specific models.
The applicability of parallel simulation capabilities of current tools is therefore limited.

This paper introduces DEVS-Ex-Machina\footnote{\url{https://bitbucket.org/bcardoen/devs-ex-machina}} (``dxex''), our simulation tool which offers multiple synchronization protocols: no synchronization (sequential execution), conservative synchronization, or optimistic synchronization.
The selected synchronization protocol is transparent to the simulated model: users should merely determine, which protocol they wish to use.
Users who simulate a wide variety of models, with different ideal synchronization protocols, can simply run the same model with different synchronization protocols.

Our tool is based on PythonPDEVS, but implemented in C++11 for increased performance.
Unlike PythonPDEVS dxex only supports multicore parallelism.

@@

The remainder of this paper is organized as follows:
Section~\ref{sec:2-background} introduces the necessary background on synchronization protocols.
Section~\ref{sec:3-features} elaborates on our design that enables this flexibility.
In Section~\ref{sec:4-performance}, we evaluate performance of our tool by comparing its different synchronization protocols, and by comparing to adevs.
Related work is discussed in Section~\ref{sec:5-related-work}.
Section~\ref{sec:6-conclusion} concludes the paper and gives future work.

### 2016-02-15 paper/2-background.tex
Spell check
Made some small changes to fit everything nicely on 8 pages

This section briefly introduces the two synchronization protocols used by dxex: conservative and optimistic synchronization.

@@

\subsection{Optimistic Synchronization}
A completely different synchronization protocol is \textit{optimistic synchronization}~\cite{TimeWarp}.
Whereas conservative synchronization prevents causality errors at all costs, optimistic synchronization allows them to happen, but corrects them.
Each node simulates as fast as possible, without taking note of any other node.
It assumes that no events occur from other nodes, unless it has explicitly received one at that time.
When this assumption is violated, the node rolls back its simulation time and state to right before the moment when the event has to be processed.
As simulation is rolled back to a time prior to the event must be processed, the event can then be processed as if no causality error ever occured.

>>>>

This section briefly introduces the synchronization protocols used by dxex: conservative and optimistic synchronization.

@@

\subsection{Optimistic Synchronization}
A completely different synchronization protocol is \textit{optimistic synchronization}~\cite{TimeWarp}.
Whereas conservative synchronization prevents causality errors at all costs, optimistic synchronization allows them to happen, but corrects them.
Each node simulates as fast as possible, without taking note of any other node.
It assumes that no events occur from other nodes, unless it has explicitly received one at that time.
When this assumption is violated, the node rolls back its simulation time and state to right before the moment when the event has to be processed.
As simulation is rolled back to a time prior to the event must be processed, the event can then be processed as if no causality error ever occurred.

### 2016-02-15 paper/4-performance.tex
Add begin/end lookahead mention.
Made some small changes to fit everything nicely on 8 pages

In this section, we evaluate performance of dxex using different synchronization protocols.
We show that the inclusion of multiple synchronization protocols does not noticeably decrease performance.
To this end, we compare to adevs, currently one of the most efficient simulation kernels~\cite{DEVSSurvey}.
CPU time and memory usage is compared for both sequential and parallel simulation.

@@

\newcommand{\modelfraction}{0.65}
\newcommand{\interfigurespace}{0.8cm}

@@

    \vspace{\interfigurespace}

@@

\subsubsection{Queue}
In the Queue model, we increase both width and depth simultaneously.
For example, the $400$ models configuration is obtained with a width and depth of $20$.
As can be seen in Figure~\ref{fig:Queue_benchmark}, dxex considerably outperforms adevs.
Through profiling, we determined that adevs spends much time on the routing of simulation messages through the hierarchy, while dxex avoids this by flattening the model structure before the simulation starts.
Both simulation tools have similar complexities, though dxex is much faster thanks to its more efficient simulation control algorithms.

@@

\subsubsection{HighInterconnect}
In the HighInterconnect model, we increase the number of atomic models, thus quadratically increasing the number of couplings and the number of external transitions.
As can be seen in Figure~\ref{fig:Interconnect_benchmark}, adevs outperforms dxex by a fair margin.
Analysis showed that this is caused by the high amount of exchanged events: event creation is much slower in dxex than it is in adevs, despite dxex's use of memory pools. For the speedup plot the amount of models is kept constant at 8.

@@

Similar to the sequential benchmarks, Figure~\ref{fig:queue_benchmark_parallel} shows that dxex outperforms adevs, but now in terms of speedup.
Results indicate that our implementation of conservative synchronization achieves much higher speedups than adevs.
This simulation is the ideal case for our conservative implementation, approaching near linear speedup, since there are no dependency cycles between models.
Conservative synchronization also seems to be better than optimistic synchronization in this case, at the cost of providing the lookahead.

@@

\subsubsection{PHold}
In the Phold model, we first investigate the influence of the fraction of remote events on the speedup.
When remote events are rare, optimistic synchronization rarely has to roll back, thus increasing performance.
With more common remote events, however, optimistic synchronization quickly slows down due to frequent rollbacks.
Conservative synchronization, on the other hand, is mostly unconcerned with the number of remote events: the mere fact that a remote event can happen, causes it to block.
Even though a single synchronization protocol is always ideal in this case, it already shows that different synchronization protocols respond differently to a changing model.
Adevs is again significantly slower during conservative synchronization.
Analysis of profiling results shows that exception handling in adevs is the main cause. 
% Addin ? To keep the models equivalent, the adevs version does not provide the \{begin,end\}Lookahead methods, which accounts for the exception handling.

@@

\vspace{1mm}
\subsubsection{Interconnect}
In the Interconnect model, we determine how broadcast communication is supported across multiple nodes.
Results are shown in Figure~\ref{fig:interconnect_benchmark_parallel}.
When the number of nodes increases, performance decreases due to increasing contention in conservative simulation and an increasing number of of rollbacks in optimistic simulation.
All models depend on each other and have no computational load whatsoever, negating any possible performance gain by executing the simulation in parallel.

@@

\subsubsection{Remarks}
Both dxex and adevs use \texttt{tcmalloc} as memory allocator, allowing for thread-local allocation.
Additionally, dxex uses memory pools to further reduce the frequency of expensive system calls (\textit{e.g.}, malloc and free).
\texttt{tcmalloc} only gradually releases memory back to the OS, whereas our pools will not do so at all.
Due to our motivation for memory usage analysis, we will only measure peak allocation.
Profiling is done using Valgrind's massif tool~\cite{Nethercote:2007:VFH:1273442.1250746}.

>>>>

In this section, we evaluate the performance of different synchronization protocols in dxex.
We also compare to adevs, currently one of the most efficient simulation kernels~\cite{DEVSSurvey}, to show that our modularity does not impede performance.
CPU time and memory usage is compared for both sequential and parallel simulation.

@@

\newcommand{\modelfraction}{0.65}

@@

@@

\subsubsection{Queue}
In the Queue model, we increase both width and depth simultaneously.
For example, the $400$ models configuration is obtained with a width and depth of $20$.
As can be seen in Figure~\ref{fig:Queue_benchmark}, dxex considerably outperforms adevs.
Through profiling, we found that adevs spends much time routing the simulation messages (\textit{e.g.}, \textit{*}, \textit{@}) through the hierarchy, whereas dxex avoids this throuhg its flattened model structure.
Both simulation tools have similar complexities, though dxex is much faster thanks to its more efficient simulation control algorithms.

@@

\subsubsection{HighInterconnect}
In the HighInterconnect model, we increase the number of atomic models, thus quadratically increasing the number of couplings and the number of external transitions.
As can be seen in Figure~\ref{fig:Interconnect_benchmark}, adevs outperforms dxex by a fair margin.
Analysis showed that this is caused by the high amount of exchanged events: event creation is much slower in dxex than it is in adevs, despite dxex's use of memory pools.

@@

Similar to the sequential benchmarks, Figure~\ref{fig:queue_benchmark_parallel} shows that dxex outperforms adevs, but now in terms of speedup.
Results indicate that our implementation of conservative synchronization achieves much higher speedups than adevs.
Approaching near linear speedup, this simulation is the ideal case for our conservative implementation, as there are no dependency cycles between models.
Conservative synchronization also seems to be better than optimistic synchronization in this case, at the cost of providing the lookahead.

@@

\subsubsection{PHold}
In the Phold model, we first investigate the influence of the fraction of remote events on the speedup.
When remote events are rare, optimistic synchronization rarely has to roll back, thus increasing performance.
With more common remote events, however, optimistic synchronization quickly slows down due to frequent rollbacks.
Conservative synchronization, on the other hand, is mostly unconcerned with the number of remote events: the mere fact that a remote event can happen, causes it to block.
Even though a single synchronization protocol is always ideal in this case, it already shows that different synchronization protocols respond differently to a changing model.
Adevs is again significantly slower during conservative synchronization.
Analysis of profiling results shows that exception handling in adevs is the main cause. 
To keep the models equivalent, the adevs version does not provide the \{begin,end\}Lookahead methods, which accounts for the exception handling.

@@

\subsubsection{Interconnect}
In the Interconnect model, we determine how broadcast communication is supported across multiple nodes.
The number of models is now kept constant at eight.
Results are shown in Figure~\ref{fig:interconnect_benchmark_parallel}.
When the number of nodes increases, performance decreases due to increasing contention in conservative simulation and an increasing number of of rollbacks in optimistic simulation.
All models depend on each other and have no computational load whatsoever, negating any possible performance gain by executing the simulation in parallel.

@@

\subsubsection{Remarks}
Both dxex and adevs use \texttt{tcmalloc} as memory allocator, allowing for thread-local allocation.
Additionally, dxex uses memory pools to further reduce the frequency of expensive system calls (\textit{e.g.}, malloc and free).
\texttt{tcmalloc} only gradually releases memory back to the OS, whereas our pools will not do so at all.
Due to our motivation for memory usage analysis, we will only measure peak allocation.
Profiling is done using Valgrind's massif tool~\cite{Nethercote:2007:VFH:1273442.1250746}.
\vspace{1cm}
